COVID-19 Policy-Compliance in Europe

### 2021-01-04 Draft.Rmd

---
title: "COVID-19 Policy-Compliance in Europe"
author: "Philipp Holz"
#date: "`r format(Sys.time(), '%d %B, %Y')`"
date: "26 December 2020"
output: 
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 6
    fig_height: 4
bibliography: References.json

@@

H\textsubscript{0}: There **is no** difference in compliance with mobility-reducing policies across European countries and this policy-compliance **does not** express a trend.

H\textsubscript{a}: There **is** a difference in compliance with mobility-reducing policies across European countries and this policy-compliance **does** express a trend.

@@

The following non-hierarchical model predicts the Mobility Index as the dependent variable, with the Stringency Index, Date and Country as the independent variables.

\begin{equation}
{Compliance Index} = {\beta}_0 + {\beta}_1 * {Date} + {\beta}_2 * {Country} + {\beta}_3 * {Publicity} + {\beta}_4 * {Cases/100.000/14 days} + {\epsilon}
\end{equation}

It yields the following result. Except for Date, none of the continuous variables Publicity and Cases/100.000/14 Days are statistically significant. Date is statistically significant at the 0.05 level. Its coefficient shows that when holding Publicity, Cases and Country constant, Compliance decreases on average by 0.07 points each day.

```{r Non-Hiearchical Model, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
## Fit the model
lm_out <- lm(data = data, ComplianceIndex ~ Date + Publicity + Date + notification_rate_per_100000_population_14.days)

@@

## Call:
## lm(formula = ComplianceIndex ~ Date + Publicity + Date + notification_rate_per_100000_population_14.days, 
##     data = data)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -39.271  -6.424  -0.826   6.354  42.727 
## 
## Coefficients:
##                                                   Estimate Std. Error t value Pr(>|t|)    
## (Intercept)                                      1.430e+03  1.181e+02  12.113   <2e-16 ***
## Date                                            -7.442e-02  6.400e-03 -11.627   <2e-16 ***
## Publicity                                       -8.653e-03  2.192e-02  -0.395    0.693    
## notification_rate_per_100000_population_14.days  6.592e-03  2.551e-03   2.584    0.010 *  
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
## 
## Residual standard error: 10.28 on 516 degrees of freedom
##   (3797 observations deleted due to missingness)
## Multiple R-squared:  0.2418,	Adjusted R-squared:  0.2374 
## F-statistic: 54.86 on 3 and 516 DF,  p-value: < 2.2e-16

@@

| Variable              | Coefficient | Std. Error | P-Value |
| --------------------- | -----------: | ----------: | :-------: |
| Date                  | −0.07       | 0.01     | ∗∗∗     |
| Publicity             | −0.01       | 0.02     |         |
| Cases/100,000/14 Days | 0.00        | 0.00     |         |
| Belgium               | −3.21       | 2.12     |         |
| Denmark               | -8.29       | 2.12       | ∗∗∗     |
| France                | -7.15       | 2.11     | ∗∗∗     |
| Germany               | −11.06      | 2.12     | ∗∗∗     |
| Ireland               | −4.07       | 2.13     |         |
| Italy                 | −8.12       | 2.17     | ∗∗∗     |
| Netherlands           | −2.97       | 2.14     |         |
| Poland                | −5.67       | 2.17     | ∗∗      |
| Spain                 | −4.79       | 2.24     | ∗       |
| Sweden                | −9.53       | 2.11     | ∗∗∗     |
| United Kingdom        | −3.25       | 2.13     |         |
R\textsuperscript{2} = 0.31;
Adj. R\textsuperscript{2} = 0.29;
Num. obs. = 520;
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.05

With Country as the categorical variable, all expressions are statistically significant at the 0.05 level, with the exception of Belgium, Ireland, the Netherlands, and the United Kingdom, when compared to Austria as the reference category. The adjusted R² indicates that the model explains 29% of the observed variance.

@@

As we can see in the regression output of our non-hierarchical model, the fixed effects estimated for the different expressions of our country variable show substantial differences, with Germany showing more than 10 points less policy-compliance than our reference country Austria. Additionally, the following plot shows that the slopes for Compliance vs. Date by Country differ slightly.

```{r Non-hierarchical different intercepts and slopes, message=FALSE, warning=FALSE, fig.align = "center"}
ggplot(data = augment(lm(data = data, 
                         ComplianceIndex ~ Date + CountryName + Publicity + notification_rate_per_100000_population_14.days)),
       aes(x = Date, y = ComplianceIndex, color = CountryName))  +
  geom_smooth(aes(y = .fitted), method = "lm", se = FALSE) +
  xlab("Month") +
  scale_x_date(breaks = date_breaks("3 months"),
               labels = date_format("%b\n%Y")) +
  scale_y_continuous(name = "Policy Compliance Index") +
  theme(
    axis.title.y = element_text(color = "black", size=12, vjust = 3),
    axis.title.x = element_text(size=12, vjust = 0.5),
    text=element_text(family="CMU Sans Serif"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(colour = "grey", linetype = "dotted"),
    plot.title = element_text(size = 11, face = "bold")) +
  ggtitle("Model Predictions for Policy Compliance in Europe Across Time")
```

Thus, to better assess the trends of policy-compliance in Europe over time, we can create a hierarchical model, allowing both random intercepts for the subgroups of Country as our grouping variable and random slopes along the Date variable.

\begin{equation}
{Compliance Index}_{ij} = {\beta}_0 + {\beta}_1 * {Date}_{ij} + {\beta}_2 * {Publicity}_{ij} + {\beta}_3 * {Cases/100,000/14 days}_{ij} + {\beta}_{1j} * {Date} + {Country}_{0j} + {\epsilon}_{0ij}
\end{equation}

@@

### Looking at the results
summary(lmer_out3)

@@

| Fixed effects         |        - |          - |       - |    -    |
| --------------------- | --------:| ----------:| -------:|:-------:|
|                       | Estimate | Std. Error | t-value | p-value |
| Date                  |    -0.07 |       0.01 |  -11.51 |   ∗∗∗   |
| Publicity             |    -0.01 |       0.02 |   -0.44 |         |
| Cases/100,000/14 Days |     0.00 |       0.00 |    1.87 |         |
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.05

As we can see, the coefficients of the fixed effects at level one have not changed substantially. Still, the fixed effect of Date is the only one that is statistically significant. The fact that the variance of the residual is almost as big as the variance of the random intercepts casts in doubt the explanatory value of our model.

@@

## Likelihood ratio test
## 
## Model 1: ComplianceIndex ~ Date + Publicity + Date + notification_rate_per_100000_population_14.days
## Model 2: ComplianceIndex ~ Date + Publicity + notification_rate_per_100000_population_14.days + 
##     (Date | CountryName)
##   #Df  LogLik Df  Chisq Pr(>Chisq)   
## 1   5 -1947.3                        
## 2   8 -1954.3  3 13.991   0.002917 **
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

@@

`Model 1: ComplianceIndex ~ Date + Publicity + Date + notification_rate_per_100000_population_14.days`

`Model 2: ComplianceIndex ~ Date + Publicity + notification_rate_per_100000_population_14.days + (Date | CountryName)`

| Model | Df  | LogLik  | Chisq  | Pr(>Chisq)  |
| ----- | --- | ------- | ------ | ----------- |
| 1     | 5   | -1947.3 |        |             |
| 2     | 8   | -1954.3 | 13.991 | 0.002917 ** |
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.05

@@

This analysis has shown that, interestingly, levels of compliance with mobility-reducing COVID-19-policies differ across selected European countries and that policy-compliance expresses an alarming, negative trend. We were able to show these findings using both a non-hierarchical and a hierarchical model. Nevertheless, this analysis has also shown that a large amount of variance in policy compliance remains unexplained by our models, which might be due to their linear character. Further attempts will have to be made in that regard.

>>>>

---
title: "COVID-19 Policy-Compliance in Europe"
author: "Philipp Holz"
#date: "`r format(Sys.time(), '%d %B, %Y')`"
date: "05 January 2021"
output: 
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 6
    fig_height: 4
bibliography: References.json

@@

H1\textsubscript{0}: There **is no** difference in compliance with mobility-reducing policies across European countries.

H1\textsubscript{a}: There **is** a difference in compliance with mobility-reducing policies across European countries.

H2\textsubscript{0}: There **is no** linear trend in such compliance over time.

H2\textsubscript{a}: There **is** a linear trend in such compliance over time.

@@

The following non-hierarchical model predicts compliance as the dependent variable, with the Stringency Index, Date and Country as the independent variables.

\begin{equation}
{Compliance} = {\beta}_0 + {\beta}_1 * {Date} + {\beta}_2 * {Country} + {\epsilon}
\end{equation}

```{r Non-Hiearchical Model 1, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
## Fit the model
lm_out <- lm(data = data, ComplianceIndex ~ Date + CountryName)

@@

## Call:
## lm(formula = ComplianceIndex ~ Date + CountryName, data = ## data)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -52.809  -6.508  -0.297   6.226  40.292 
## 
## Coefficients:
##                             Estimate Std. Error t value ## Pr(>|t|)    
## (Intercept)                1.223e+03  3.575e+01  34.211  < ## 2e-16 ***
## Date                      -6.291e-02  1.936e-03 -32.489  < ## 2e-16 ***
## CountryNameBelgium        -3.599e+00  8.334e-01  -4.318 ## 1.61e-05 ***
## CountryNameDenmark        -9.119e+00  8.314e-01 -10.969  < ## 2e-16 ***
## CountryNameFrance         -7.065e+00  8.314e-01  -8.497  < ## 2e-16 ***
## CountryNameGermany        -1.196e+01  8.314e-01 -14.385  < ## 2e-16 ***
## CountryNameIreland        -5.587e+00  8.341e-01  -6.698 ## 2.43e-11 ***
## CountryNameItaly          -7.642e+00  8.348e-01  -9.154  < ## 2e-16 ***
## CountryNameNetherlands    -5.099e+00  8.444e-01  -6.039 ## 1.70e-09 ***
## CountryNamePoland         -6.166e+00  8.327e-01  -7.404 ## 1.63e-13 ***
## CountryNameSpain          -4.647e+00  8.348e-01  -5.566 ## 2.79e-08 ***
## CountryNameSweden         -1.021e+01  8.314e-01 -12.279  < ## 2e-16 ***
## CountryNameUnited Kingdom -2.807e+00  8.314e-01  -3.376 ## 0.000742 ***
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
## 
## Residual standard error: 10.32 on 3646 degrees of freedom
##   (670 observations deleted due to missingness)
## Multiple R-squared:  0.2791,	Adjusted R-squared:  0.2768 
## F-statistic: 117.7 on 12 and 3646 DF,  p-value: < 2.2e-16
```

It yields the following result.

| Variable                  | Coefficient | Std. Error | P-Value |
| ------------------------- | ----------- | ---------- | ------- |
| Date                      | -6.291e-02  | 1.936e-03  | \*\*\*  |
| Belgium                   | -3.599e+00  | 8.334e-01  | \*\*\*  |
| Denmark                   | -9.119e+00  | 8.314e-01  | \*\*\*  |
| France                    | -7.065e+00  | 8.314e-01  | \*\*\*  |
| Germany                   | -1.196e+01  | 8.314e-01  | \*\*\*  |
| Ireland                   | -5.587e+00  | 8.341e-01  | \*\*\*  |
| Italy                     | -7.642e+00  | 8.348e-01  | \*\*\*  |
| Netherlands               | -5.099e+00  | 8.444e-01  | \*\*\*  |
| Poland                    | -6.166e+00  | 8.327e-01  | \*\*\*  |
| Spain                     | -4.647e+00  | 8.348e-01  | \*\*\*  |
| Sweden                    | -1.021e+01  | 8.314e-01  | \*\*\*  |
| United Kingdom            | -2.807e+00  | 8.314e-01  | \*\*\*  |
R\textsuperscript{2} = 0.28;
Adj. R\textsuperscript{2} = 0.28;
Num. obs. = 520;
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.0

The coefficients for Data and the individual Countries, when compared to Austria, are statistically significant. The coefficient for date is negative, with only little standard error. Also the coefficients for the countries are negative, signalling that Austria showed the highest level of compliance, with Germany and Sweden at the bottom of the list. We can control this relationship for the influence of public attention to the pandemic and for COVID-19 cases per 100.000 and 14 days.

\begin{equation}
{Compliance} = {\beta}_0 + {\beta}_1 * {Date} + {\beta}_2 * {Country} + {\beta}_3 * {Public Attention} + {\beta}_4 * {Cases/(100.000 * 14 days)} + {\epsilon}
\end{equation}

```{r Non-Hiearchical Model 2, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
## Fit the model
lm_out <- lm(data = data, ComplianceIndex ~ Date + CountryName + Publicity + notification_rate_per_100000_population_14.days)

## Call:
## lm(formula = ComplianceIndex ~ Date + CountryName + Publicity + 
##     notification_rate_per_100000_population_14.days, data = data)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -36.552  -6.433  -0.435   5.822  37.089 
## 
## Coefficients:
##                                                   Estimate Std. Error t value Pr(>|t|)    
## (Intercept)                                      1.381e+03  1.155e+02  11.953  < 2e-16 ***
## Date                                            -7.142e-02  6.264e-03 -11.401  < 2e-16 ***
## CountryNameBelgium                              -3.211e+00  2.122e+00  -1.514 0.130771    
## CountryNameDenmark                              -8.285e+00  2.124e+00  -3.902 0.000108 ***
## CountryNameFrance                               -7.145e+00  2.112e+00  -3.383 0.000773 ***
## CountryNameGermany                              -1.106e+01  2.124e+00  -5.209 2.77e-07 ***
## CountryNameIreland                              -4.066e+00  2.133e+00  -1.906 0.057210 .  
## CountryNameItaly                                -8.116e+00  2.165e+00  -3.748 0.000199 ***
## CountryNameNetherlands                          -2.967e+00  2.142e+00  -1.385 0.166606    
## CountryNamePoland                               -5.667e+00  2.169e+00  -2.613 0.009253 ** 
## CountryNameSpain                                -4.790e+00  2.239e+00  -2.139 0.032904 *  
## CountryNameSweden                               -9.533e+00  2.113e+00  -4.511 8.03e-06 ***
## CountryNameUnited Kingdom                       -3.250e+00  2.126e+00  -1.529 0.126923    
## Publicity                                       -1.085e-02  2.393e-02  -0.453 0.650518    
## notification_rate_per_100000_population_14.days  4.710e-03  2.557e-03   1.842 0.066033 .  
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
## 
## Residual standard error: 9.907 on 505 degrees of freedom
##   (3809 observations deleted due to missingness)
## Multiple R-squared:  0.3102,	Adjusted R-squared:  0.2911 
## F-statistic: 16.22 on 14 and 505 DF,  p-value: < 2.2e-16

@@

It yields the following result. Except for Date, none of the continuous variables Public Attention and Cases per 100,000 and 14 Days are statistically significant. Date is statistically significant at the 0.05 level. Its coefficient shows that when holding Publicity, Cases and Country constant, Compliance decreases on average by 0.07 points each day. Interestingly, some of the coefficients for individual countries are not significant anymore when controlling for public attention and COVID-19 cases. In the following, a random intercept will be included for country, in order to improve model fit.

 | Variable                      | Coefficient | Std. Error | P-Value |
 | ----------------------------- | ----------- | ---------- | ------- |
 | Date                          | -7.142e-02  | 6.264e-03  | \*\*\*  |
 | Belgium                       | -3.211e+00  | 2.122e+00  |         |
 | Denmark                       | -8.285e+00  | 2.124e+00  | \*\*\*  |
 | France                        | -7.145e+00  | 2.112e+00  | \*\*\*  |
 | Germany                       | -1.106e+01  | 2.124e+00  | \*\*\*  |
 | Ireland                       | -4.066e+00  | 2.133e+00  | .       |
 | Italy                         | -8.116e+00  | 2.165e+00  | \*\*\*  |
 | Netherlands                   | -2.967e+00  | 2.142e+00  |         |
 | Poland                        | -5.667e+00  | 2.169e+00  | \*\*    |
 | Spain                         | -4.790e+00  | 2.239e+00  | \*      |
 | Sweden                        | -9.533e+00  | 2.113e+00  | \*\*\*  |
 | United Kingdom                | -3.250e+00  | 2.126e+00  |         |
 | Public Attention              | -1.085e-02  | 2.393e-02  |         |
 | Cases per 100,000 and 14 Days | 4.710e-03   | 2.557e-03  | .       |
R\textsuperscript{2} = 0.31;
Adj. R\textsuperscript{2} = 0.29;
Num. obs. = 520;
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.05

@@

As we can see in the regression output of our non-hierarchical model, the fixed effects estimated for the different expressions of our country variable show substantial differences, with Germany showing more than 10 points less policy-compliance than our reference country Austria.

To better assess the trends of policy-compliance in Europe over time, we can create a hierarchical model, allowing both random intercepts for the subgroups of Country as our grouping variable and random slopes along the Date variable.

\begin{equation}
{Compliance}_{ij} = {\beta}_0 + {\beta}_1 * {Date}_{ij} + {\beta}_2 * {Public Attention}_{ij} + {\beta}_3 * {Cases/(100,000 * 14 days)}_{ij} + {\beta}_{1j} * {Date} + {Country}_{0j} + {\epsilon}_{0ij}
\end{equation}

@@

### Looking at the results
summary(lmer_out)

@@

| Fixed effects                 |        - |          - |       - |    -    |
| ------------------------------| --------:| ----------:| -------:|:-------:|
|                               | Estimate | Std. Error | t-value | p-value |
| Date                          |    -0.07 |       0.01 |  -11.51 |   ∗∗∗   |
| Public Attention              |    -0.01 |       0.02 |   -0.44 |         |
| Cases per 100,000 and 14 Days |     0.00 |       0.00 |    1.87 |         |
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.05

As we can see, the coefficients of the fixed effects at level one have not changed substantially. Still, the fixed effect of Date is the only one that is statistically significant.

The following plot shows the rate at which compliance changes per day in different European countries, according to our hierarchical model, that controls for public attention and COVID-19-cases.

```{r slope of compliance-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center"}

# Model
lmer_out <- lmer(data = data, ComplianceIndex ~ Date + Publicity + notification_rate_per_100000_population_14.days +  (Date | CountryName))

## Extracting out the fixed-effect slope for Date
Date_slope <- fixef(lmer_out)['Date']

## Extracting out the random-effect slopes for Country
Country_slope <- ranef(lmer_out)$CountryName

## Creating a new column for the slope
Country_slope$slope <- Country_slope$Date + Date_slope

## Using row names to create a county name column
Country_slope$country <- rownames(Country_slope)

## Ordered county-level factor based upon slope values
Country_slope$country_plot <- factor(Country_slope$country, 
                                   levels = Country_slope$country[order(Country_slope$slope)])

## Plotting the results
ggplot(data = Country_slope, aes(x = country_plot, y = slope)) + 
  geom_point() +
  coord_flip() + 
  theme_bw() + 
  ylab("Change in Compliance per Day")  +
  xlab("Country") +
  theme(
    axis.title.y = element_text(color = "black", size=12, vjust = 3),
    axis.title.x = element_text(size=12, vjust = 0.5),
    text=element_text(family="CMU Sans Serif"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(colour = "grey", linetype = "dotted"),
    plot.title = element_text(size = 11, face = "bold")) +
  ggtitle("Rate of Change in Compliance in Europe")
```

@@

## Likelihood ratio test
## 
## Model 1: ComplianceIndex ~ Date + CountryName + Publicity + notification_rate_per_100000_population_14.days
## Model 2: ComplianceIndex ~ Date + Publicity + notification_rate_per_100000_population_14.days + 
##     (Date | CountryName)
##   #Df  LogLik Df Chisq Pr(>Chisq)    
## 1  16 -1922.8                        
## 2   8 -1954.3 -8 63.15  1.119e-10 ***
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

@@

`Model 1: ComplianceIndex ~ Date + Country + Public Attention + Cases per 100,000 and 14 Days`

`Model 2: ComplianceIndex ~ Date + Public Attention + Cases per 100,000 and 14 Days + (Date | Country)`

| Model | Df  | LogLik  | Chisq  | P-Value       |
| ----- | --- | ------- | ------ | --------------|
| 1     | 16  | -1922.8 |        |               |
| 2     | 8   | -1954.3 | 63.15  | 1.119e-10 *** |
∗∗∗ p < 0.001; ∗∗ p < 0.01; ∗ p < 0.05

@@

This analysis has shown that, interestingly, levels of compliance with mobility-reducing COVID-19-policies differ across selected European countries and that policy-compliance expresses an alarming, negative trend. Additionally, we were able to show how the *rate* at which policy-compliance decreases differs across European countries.
